---
title: "Literature Data Integration Source Progress Report"
author: "BIEN Literature_Data_To_BIENdb"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(readr)
library(knitr)
library(DT)
library(leaflet)

project_root <- getwd()
intake_file <- file.path(project_root, "data", "occurrence_source_intake.csv")
occurrence_dir <- file.path(project_root, "data", "occurrences")

if (!file.exists(intake_file)) {
  stop("Missing required file: data/occurrence_source_intake.csv")
}

intake <- readr::read_csv(intake_file, show_col_types = FALSE)

required_cols <- c(
  "source_id", "display_name", "intake_type", "source_group",
  "canonical_url_or_doi", "harvest_status", "next_action", "notes"
)
missing_cols <- setdiff(required_cols, names(intake))
if (length(missing_cols) > 0) {
  stop(paste("Missing required columns in intake CSV:", paste(missing_cols, collapse = ", ")))
}

compiled_paths <- file.path(occurrence_dir, intake$source_id, "compiled_occurrences.csv")
compiled_exists <- file.exists(compiled_paths)

read_source_summary <- function(p) {
  tryCatch({
    df <- readr::read_csv(p, show_col_types = FALSE)
    total <- nrow(df)
    georef <- if (all(c("decimalLatitude", "decimalLongitude") %in% names(df))) {
      sum(!is.na(df$decimalLatitude) & !is.na(df$decimalLongitude) &
          df$decimalLatitude != "" & df$decimalLongitude != "", na.rm = TRUE)
    } else { 0L }
    binomial <- if ("species" %in% names(df)) {
      sum(!is.na(df$species) & df$species != "" &
          grepl(" ", trimws(as.character(df$species))), na.rm = TRUE)
    } else { 0L }
    list(total = total, georef = as.integer(georef), binomial = as.integer(binomial), data = df)
  }, error = function(e) {
    list(total = NA_integer_, georef = NA_integer_, binomial = NA_integer_, data = NULL)
  })
}

source_summaries <- vector("list", length(compiled_paths))
for (i in seq_along(compiled_paths)) {
  if (compiled_exists[i]) {
    source_summaries[[i]] <- read_source_summary(compiled_paths[i])
  } else {
    source_summaries[[i]] <- list(total = NA_integer_, georef = NA_integer_, binomial = NA_integer_, data = NULL)
  }
}

compiled_rows     <- sapply(source_summaries, function(s) s$total)
compiled_georef   <- sapply(source_summaries, function(s) s$georef)
compiled_binomial <- sapply(source_summaries, function(s) s$binomial)

intake_enriched <- intake %>%
  mutate(
    canonical_url_or_doi = ifelse(is.na(canonical_url_or_doi) | canonical_url_or_doi == "", "unknown/unverified", canonical_url_or_doi),
    next_action = ifelse(is.na(next_action) | next_action == "", "unknown/unverified", next_action),
    notes = ifelse(is.na(notes) | notes == "", "unknown/unverified", notes),
    compiled_file_exists = ifelse(compiled_exists, "yes", "no"),
    compiled_row_count   = compiled_rows,
    georef_records       = compiled_georef,
    binomial_records     = compiled_binomial,
    pct_georef   = ifelse(!is.na(compiled_rows) & compiled_rows > 0,
                          round(compiled_georef   / compiled_rows * 100, 1), NA_real_),
    pct_binomial = ifelse(!is.na(compiled_rows) & compiled_rows > 0,
                          round(compiled_binomial / compiled_rows * 100, 1), NA_real_),
    status_bucket = case_when(
      harvest_status == "compiled" ~ "accessed_or_ingested",
      harvest_status %in% c("pending_review", "pending_manual_access") ~ "pending",
      TRUE ~ "pending"
    )
  )

# Combined georeferenced points for leaflet map (capped at 50,000 for performance)
all_georef_pts <- do.call(rbind, lapply(seq_along(source_summaries), function(i) {
  s <- source_summaries[[i]]
  if (is.null(s$data)) return(NULL)
  df <- s$data
  if (!all(c("decimalLatitude", "decimalLongitude") %in% names(df))) return(NULL)
  df <- df[!is.na(df$decimalLatitude) & !is.na(df$decimalLongitude) &
           df$decimalLatitude != "" & df$decimalLongitude != "", ]
  if (nrow(df) == 0) return(NULL)
  data.frame(
    source_id = intake$source_id[i],
    species   = if ("species" %in% names(df)) as.character(df$species) else NA_character_,
    lat = suppressWarnings(as.numeric(df$decimalLatitude)),
    lon = suppressWarnings(as.numeric(df$decimalLongitude)),
    stringsAsFactors = FALSE
  )
}))
all_georef_pts <- all_georef_pts[!is.na(all_georef_pts$lat) & !is.na(all_georef_pts$lon), ]
if (!is.null(all_georef_pts) && nrow(all_georef_pts) > 50000) {
  set.seed(42)
  all_georef_pts <- all_georef_pts[sample(nrow(all_georef_pts), 50000), ]
}

accessed_tbl <- intake_enriched %>%
  filter(status_bucket == "accessed_or_ingested") %>%
  arrange(source_group, source_id)

pending_tbl <- intake_enriched %>%
  filter(status_bucket == "pending") %>%
  arrange(harvest_status, source_group, source_id)

status_counts <- intake_enriched %>%
  count(harvest_status, name = "n") %>%
  arrange(desc(n))

n_total <- nrow(intake_enriched)
n_accessed <- nrow(accessed_tbl)
n_pending <- nrow(pending_tbl)
pct_accessed <- ifelse(n_total > 0, round(100 * n_accessed / n_total, 1), NA_real_)
pct_pending <- ifelse(n_total > 0, round(100 * n_pending / n_total, 1), NA_real_)

what_done <- intake_enriched %>%
  filter(status_bucket == "accessed_or_ingested") %>%
  count(source_group, name = "sources_completed") %>%
  arrange(desc(sources_completed), source_group)

pending_actions <- pending_tbl %>%
  transmute(source_id, display_name, harvest_status, next_action)

# ── Literature papers setup ─────────────────────────────────────────────────
papers_cfg_file <- file.path(project_root, "config", "papers.csv")
papers_cfg <- tryCatch(
  readr::read_csv(papers_cfg_file, show_col_types = FALSE),
  error = function(e) NULL
)

if (!is.null(papers_cfg)) {
  processed_dir <- file.path(project_root, "data", "processed")

  summary_list <- lapply(papers_cfg$paper_id, function(pid) {
    f <- file.path(processed_dir, paste0(pid, "_staging_summary.csv"))
    if (file.exists(f)) {
      tryCatch(readr::read_csv(f, show_col_types = FALSE),
               error = function(e) NULL)
    } else {
      NULL
    }
  })
  has_summary <- !sapply(summary_list, is.null)

  papers_summary_raw <- if (any(has_summary)) {
    do.call(rbind, summary_list[has_summary])
  } else {
    NULL
  }

  summary_df_for_join <- if (!is.null(papers_summary_raw)) {
    papers_summary_raw
  } else {
    data.frame(
      paper_id = character(0), n_rows = integer(0),
      n_with_coords = integer(0), n_with_country = integer(0),
      n_with_state_province = integer(0), n_with_county_parish = integer(0),
      n_with_municipality = integer(0), n_habit_trait_rows = integer(0),
      n_unique_species = integer(0), generated_at_utc = character(0),
      stringsAsFactors = FALSE
    )
  }

  papers_all <- papers_cfg %>%
    dplyr::left_join(summary_df_for_join, by = "paper_id") %>%
    dplyr::mutate(
      status     = ifelse(!is.na(n_rows), "Processed", "Pending"),
      pct_georef = ifelse(!is.na(n_rows) & n_rows > 0,
                         round(n_with_coords / n_rows * 100, 1), NA_real_)
    )

  # Load georeferenced points from bien_staging.csv files
  lit_georef_list <- lapply(papers_cfg$paper_id, function(pid) {
    f <- file.path(processed_dir, paste0(pid, "_bien_staging.csv"))
    if (!file.exists(f)) return(NULL)
    tryCatch({
      df <- readr::read_csv(f, show_col_types = FALSE)
      if (!all(c("latitude", "longitude") %in% names(df))) return(NULL)
      df <- df[!is.na(df$latitude) & !is.na(df$longitude) &
               df$latitude != "" & df$longitude != "", ]
      if (nrow(df) == 0) return(NULL)
      data.frame(
        paper_id       = pid,
        latitude       = suppressWarnings(as.numeric(df$latitude)),
        longitude      = suppressWarnings(as.numeric(df$longitude)),
        name_submitted = if ("name_submitted" %in% names(df))
                           as.character(df$name_submitted)
                         else NA_character_,
        stringsAsFactors = FALSE
      )
    }, error = function(e) NULL)
  })

  lit_georef_pts <- do.call(rbind, Filter(Negate(is.null), lit_georef_list))
  if (!is.null(lit_georef_pts) && nrow(lit_georef_pts) > 0) {
    lit_georef_pts <- lit_georef_pts[
      !is.na(lit_georef_pts$latitude) & !is.na(lit_georef_pts$longitude), ]
    if (nrow(lit_georef_pts) > 50000) {
      set.seed(42)
      lit_georef_pts <- lit_georef_pts[sample(nrow(lit_georef_pts), 50000), ]
    }
  } else {
    lit_georef_pts <- NULL
  }

  n_papers_total      <- nrow(papers_cfg)
  n_papers_processed  <- sum(has_summary)
  n_papers_pending    <- n_papers_total - n_papers_processed
  n_staging_rows_total <- if (!is.null(papers_summary_raw))
                            sum(papers_summary_raw$n_rows, na.rm = TRUE)
                          else 0L
  n_lit_georef_total  <- if (!is.null(lit_georef_pts)) nrow(lit_georef_pts) else 0L

} else {
  papers_all          <- NULL
  lit_georef_pts      <- NULL
  n_papers_total      <- 0L
  n_papers_processed  <- 0L
  n_papers_pending    <- 0L
  n_staging_rows_total <- 0L
  n_lit_georef_total  <- 0L
}

# ── Consolidated map: combine literature papers + occurrence intake ───────────
# Normalise lit_georef_pts (latitude/longitude/paper_id/name_submitted)
lit_pts_norm <- if (!is.null(lit_georef_pts) && nrow(lit_georef_pts) > 0) {
  data.frame(
    source_label = lit_georef_pts$paper_id,
    lat          = as.numeric(lit_georef_pts$latitude),
    lon          = as.numeric(lit_georef_pts$longitude),
    label        = lit_georef_pts$name_submitted,
    track        = "Literature Paper",
    stringsAsFactors = FALSE
  )
} else NULL

# Normalise all_georef_pts (lat/lon/source_id/species)
occ_pts_norm <- if (!is.null(all_georef_pts) && nrow(all_georef_pts) > 0) {
  data.frame(
    source_label = all_georef_pts$source_id,
    lat          = as.numeric(all_georef_pts$lat),
    lon          = as.numeric(all_georef_pts$lon),
    label        = all_georef_pts$species,
    track        = "Occurrence Intake",
    stringsAsFactors = FALSE
  )
} else NULL

combined_georef_pts <- do.call(rbind, Filter(Negate(is.null), list(lit_pts_norm, occ_pts_norm)))
if (!is.null(combined_georef_pts) && nrow(combined_georef_pts) > 0) {
  combined_georef_pts <- combined_georef_pts[
    !is.na(combined_georef_pts$lat) & !is.na(combined_georef_pts$lon), ]
  if (nrow(combined_georef_pts) > 50000) {
    set.seed(42)
    combined_georef_pts <- combined_georef_pts[sample(nrow(combined_georef_pts), 50000), ]
  }
} else {
  combined_georef_pts <- NULL
}
n_combined_georef <- if (!is.null(combined_georef_pts)) nrow(combined_georef_pts) else 0L
```

# Literature Paper Integration Status

## Executive Sub-Summary

- Total papers configured: **`r n_papers_total`**
- Papers processed/staged: **`r n_papers_processed`**
- Papers pending: **`r n_papers_pending`**
- Total staging rows across all processed papers: **`r n_staging_rows_total`**
- Total georeferenced observations (literature papers): **`r n_lit_georef_total`**

## Per-Paper Status Table

```{r lit-papers-table}
if (!is.null(papers_all)) {
  papers_display <- papers_all %>%
    dplyr::mutate(
      citation_short = substr(citation, 1, 80),
      doi_link = ifelse(
        !is.na(doi) & doi != "",
        sprintf('<a href="https://doi.org/%s" target="_blank">%s</a>', doi, doi),
        "—"
      )
    ) %>%
    dplyr::select(
      paper_id,
      citation_short,
      year,
      journal,
      doi_link,
      status,
      total_records    = n_rows,
      georef_records   = n_with_coords,
      pct_georef,
      unique_species   = n_unique_species,
      habit_trait_rows = n_habit_trait_rows
    )

  DT::datatable(
    papers_display,
    escape     = FALSE,
    rownames   = FALSE,
    options    = list(pageLength = 15, scrollX = TRUE),
    caption    = "All configured literature papers with staging status and quality metrics. Pending papers show NA for numeric fields.",
    colnames   = c(
      "Paper ID", "Citation (truncated)", "Year", "Journal", "DOI",
      "Status", "Total Records", "Georef Records", "% Georef",
      "Unique Species", "Habit/Trait Rows"
    )
  )
} else {
  cat("config/papers.csv not found — literature paper section unavailable.")
}
```

## Geographic Coverage — All Sources

All georeferenced points from both literature papers and occurrence intake sources are shown together in the consolidated map below (Section 3.2).

# Executive Summary

- Report timestamp: `r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`
- Source registry file: `data/occurrence_source_intake.csv`
- Total sources tracked: **`r n_total`**
- Sources accessed/ingested (`harvest_status == "compiled"`): **`r n_accessed`** (`r pct_accessed`%)
- Sources pending (`harvest_status != "compiled"`): **`r n_pending`** (`r pct_pending`%)

```{r status-breakdown}
knitr::kable(status_counts, caption = "Harvest status breakdown from intake registry")
```

# Sources Already Accessed/Ingested

Evidence columns are included directly from repository artifacts. If evidence is unavailable, fields are marked as `unknown/unverified`.

```{r accessed-table}
accessed_display <- accessed_tbl %>%
  select(
    source_id,
    display_name,
    intake_type,
    source_group,
    canonical_url_or_doi,
    harvest_status,
    compiled_file_exists,
    compiled_row_count,
    notes
  )

DT::datatable(
  accessed_display,
  options = list(pageLength = 15, scrollX = TRUE),
  rownames = FALSE,
  caption = "Sources with compiled status"
)
```

# Compiled Source Quality Summary

Per-source quality metrics computed from compiled occurrence CSVs at render time.

```{r quality-summary-table}
quality_tbl <- intake_enriched %>%
  filter(harvest_status == "compiled") %>%
  select(source_id, display_name,
         total_records    = compiled_row_count,
         georef_records,  pct_georef,
         binomial_records, pct_binomial) %>%
  arrange(source_id)

totals_row <- data.frame(
  source_id        = "TOTAL",
  display_name     = "",
  total_records    = sum(quality_tbl$total_records,    na.rm = TRUE),
  georef_records   = sum(quality_tbl$georef_records,   na.rm = TRUE),
  pct_georef       = round(sum(quality_tbl$georef_records,   na.rm = TRUE) /
                           max(sum(quality_tbl$total_records, na.rm = TRUE), 1) * 100, 1),
  binomial_records = sum(quality_tbl$binomial_records, na.rm = TRUE),
  pct_binomial     = round(sum(quality_tbl$binomial_records, na.rm = TRUE) /
                           max(sum(quality_tbl$total_records, na.rm = TRUE), 1) * 100, 1),
  stringsAsFactors = FALSE
)

quality_display <- rbind(quality_tbl, totals_row)

knitr::kable(
  quality_display,
  col.names = c("Source ID", "Display Name", "Total Records",
                "Georef Records", "% Georef",
                "Binomial Records", "% Binomial"),
  caption = "Per-source quality metrics for compiled occurrence files"
)
```

## Consolidated Geographic Map — All Sources

All **`r n_combined_georef`** georeferenced points across literature papers and compiled occurrence intake sources (sampled to 50,000 max for performance). Points are coloured by source; shape of popup shows data track (Literature Paper vs Occurrence Intake).

```{r consolidated-map}
if (!is.null(combined_georef_pts) && nrow(combined_georef_pts) > 0) {
  all_labels  <- unique(combined_georef_pts$source_label)
  n_labels    <- length(all_labels)
  pal_colors  <- if (n_labels <= 12) {
    RColorBrewer::brewer.pal(max(3L, n_labels), "Paired")[seq_len(n_labels)]
  } else {
    colorRampPalette(RColorBrewer::brewer.pal(12, "Paired"))(n_labels)
  }
  combined_pal <- colorFactor(palette = pal_colors, domain = all_labels)

  leaflet(combined_georef_pts, options = leafletOptions(minZoom = 1)) %>%
    addProviderTiles("CartoDB.Positron") %>%
    addCircleMarkers(
      lng         = ~lon,
      lat         = ~lat,
      color       = ~combined_pal(source_label),
      radius      = 3,
      stroke      = FALSE,
      fillOpacity = 0.65,
      popup       = ~paste0(
        "<b>", track, ":</b> ", source_label,
        "<br><b>Species:</b> ", label
      )
    ) %>%
    addLegend(
      position = "bottomright",
      pal      = combined_pal,
      values   = ~source_label,
      title    = "Source",
      opacity  = 0.85
    )
} else {
  cat("No georeferenced points available to map.")
}
```

# Sources Pending (Waiting List)

```{r pending-table}
pending_display <- pending_tbl %>%
  select(
    source_id,
    display_name,
    intake_type,
    source_group,
    canonical_url_or_doi,
    harvest_status,
    next_action,
    notes
  )

DT::datatable(
  pending_display,
  options = list(pageLength = 20, scrollX = TRUE),
  rownames = FALSE,
  caption = "Sources pending review or manual access"
)
```

# What Has Been Done

```{r done-summary}
knitr::kable(what_done, caption = "Completed (compiled) sources by source_group")
```

- The registry has explicit status entries for all listed sources via `harvest_status`.
- Compiled-source directories were checked under `data/occurrences/<source_id>/compiled_occurrences.csv` for evidence.
- Canonical links/DOIs, action notes, and source notes were pulled from the intake registry without manual reinterpretation.

# What Remains To Be Done

Actionable checklist generated from pending rows:

```{r pending-actions-table}
DT::datatable(
  pending_actions,
  options = list(pageLength = 20, scrollX = TRUE),
  rownames = FALSE,
  caption = "Pending source action checklist"
)
```

- [ ] Resolve all `pending_manual_access` sources with direct coordinator contact and access confirmation.
- [ ] Execute each pending source `next_action` and capture outputs in `data/occurrences/<source_id>/`.
- [ ] Update `harvest_status` to `compiled` only after compiled output is written and validated.
- [ ] Re-render this report after each status update to keep the waiting list current.

# Reproducibility Notes

- This report is fully generated from repository files at render time.
- No source status was inferred from external systems.
- Missing or blank fields are reported as `unknown/unverified`.
- Render command:

```bash
cd /Users/brianjenquist/VSCode/Literature_Data_To_BIENdb
Rscript -e "rmarkdown::render('source_data_progress_report.Rmd')"
```
