---
title: "Literature Data Integration Source Progress Report"
author: "BIEN Literature_Data_To_BIENdb"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(readr)
library(tidyr)
library(knitr)
library(DT)
library(leaflet)
library(ggplot2)

project_root <- getwd()
intake_file <- file.path(project_root, "data", "occurrence_source_intake.csv")
occurrence_dir <- file.path(project_root, "data", "occurrences")

if (!file.exists(intake_file)) {
  stop("Missing required file: data/occurrence_source_intake.csv")
}

intake <- readr::read_csv(intake_file, show_col_types = FALSE)

required_cols <- c(
  "source_id", "display_name", "intake_type", "source_group",
  "canonical_url_or_doi", "harvest_status", "next_action", "notes"
)
missing_cols <- setdiff(required_cols, names(intake))
if (length(missing_cols) > 0) {
  stop(paste("Missing required columns in intake CSV:", paste(missing_cols, collapse = ", ")))
}

compiled_paths <- file.path(occurrence_dir, intake$source_id, "compiled_occurrences.csv")
compiled_exists <- file.exists(compiled_paths)

read_source_summary <- function(p) {
  tryCatch({
    df <- readr::read_csv(p, show_col_types = FALSE)
    total <- nrow(df)
    georef <- if (all(c("decimalLatitude", "decimalLongitude") %in% names(df))) {
      sum(!is.na(df$decimalLatitude) & !is.na(df$decimalLongitude) &
          df$decimalLatitude != "" & df$decimalLongitude != "", na.rm = TRUE)
    } else { 0L }
    binomial <- if ("species" %in% names(df)) {
      sum(!is.na(df$species) & df$species != "" &
          grepl(" ", trimws(as.character(df$species))), na.rm = TRUE)
    } else { 0L }
    list(total = total, georef = as.integer(georef), binomial = as.integer(binomial), data = df)
  }, error = function(e) {
    list(total = NA_integer_, georef = NA_integer_, binomial = NA_integer_, data = NULL)
  })
}

source_summaries <- vector("list", length(compiled_paths))
for (i in seq_along(compiled_paths)) {
  if (compiled_exists[i]) {
    source_summaries[[i]] <- read_source_summary(compiled_paths[i])
  } else {
    source_summaries[[i]] <- list(total = NA_integer_, georef = NA_integer_, binomial = NA_integer_, data = NULL)
  }
}

compiled_rows     <- sapply(source_summaries, function(s) s$total)
compiled_georef   <- sapply(source_summaries, function(s) s$georef)
compiled_binomial <- sapply(source_summaries, function(s) s$binomial)

intake_enriched <- intake %>%
  mutate(
    canonical_url_or_doi = ifelse(is.na(canonical_url_or_doi) | canonical_url_or_doi == "", "unknown/unverified", canonical_url_or_doi),
    next_action = ifelse(is.na(next_action) | next_action == "", "unknown/unverified", next_action),
    notes = ifelse(is.na(notes) | notes == "", "unknown/unverified", notes),
    compiled_file_exists = ifelse(compiled_exists, "yes", "no"),
    compiled_row_count   = compiled_rows,
    georef_records       = compiled_georef,
    binomial_records     = compiled_binomial,
    pct_georef   = ifelse(!is.na(compiled_rows) & compiled_rows > 0,
                          round(compiled_georef   / compiled_rows * 100, 1), NA_real_),
    pct_binomial = ifelse(!is.na(compiled_rows) & compiled_rows > 0,
                          round(compiled_binomial / compiled_rows * 100, 1), NA_real_),
    status_bucket = case_when(
      harvest_status == "compiled" ~ "accessed_or_ingested",
      harvest_status %in% c("pending_review", "pending_manual_access") ~ "pending",
      TRUE ~ "pending"
    )
  )

# Combined georeferenced points for leaflet map (capped at 50,000 for performance)
all_georef_pts <- do.call(rbind, lapply(seq_along(source_summaries), function(i) {
  s <- source_summaries[[i]]
  if (is.null(s$data)) return(NULL)
  df <- s$data
  if (!all(c("decimalLatitude", "decimalLongitude") %in% names(df))) return(NULL)
  df <- df[!is.na(df$decimalLatitude) & !is.na(df$decimalLongitude) &
           df$decimalLatitude != "" & df$decimalLongitude != "", ]
  if (nrow(df) == 0) return(NULL)
  data.frame(
    source_id = intake$source_id[i],
    species   = if ("species" %in% names(df)) as.character(df$species) else NA_character_,
    lat = suppressWarnings(as.numeric(df$decimalLatitude)),
    lon = suppressWarnings(as.numeric(df$decimalLongitude)),
    stringsAsFactors = FALSE
  )
}))
all_georef_pts <- all_georef_pts[!is.na(all_georef_pts$lat) & !is.na(all_georef_pts$lon), ]
if (!is.null(all_georef_pts) && nrow(all_georef_pts) > 50000) {
  set.seed(42)
  all_georef_pts <- all_georef_pts[sample(nrow(all_georef_pts), 50000), ]
}

accessed_tbl <- intake_enriched %>%
  filter(status_bucket == "accessed_or_ingested") %>%
  arrange(source_group, source_id)

pending_tbl <- intake_enriched %>%
  filter(status_bucket == "pending") %>%
  arrange(harvest_status, source_group, source_id)

status_counts <- intake_enriched %>%
  count(harvest_status, name = "n") %>%
  arrange(desc(n))

n_total <- nrow(intake_enriched)
n_accessed <- nrow(accessed_tbl)
n_pending <- nrow(pending_tbl)
pct_accessed <- ifelse(n_total > 0, round(100 * n_accessed / n_total, 1), NA_real_)
pct_pending <- ifelse(n_total > 0, round(100 * n_pending / n_total, 1), NA_real_)

recent_intake_sources <- c("manual_ukrainian_kurgans_gbif", "manual_ecat_central_africa")
recent_additions <- intake_enriched %>%
  filter(source_id %in% recent_intake_sources) %>%
  mutate(
    note_summary = sprintf(
      "%s — %s compiled rows, %s georeferenced",
      display_name,
      ifelse(is.na(compiled_row_count), "NA", format(compiled_row_count, big.mark = ",")),
      ifelse(is.na(georef_records), "NA", format(georef_records, big.mark = ","))
    )
  )

what_done <- intake_enriched %>%
  filter(status_bucket == "accessed_or_ingested") %>%
  count(source_group, name = "sources_completed") %>%
  arrange(desc(sources_completed), source_group)

pending_actions <- pending_tbl %>%
  transmute(source_id, display_name, harvest_status, next_action)

# ── Literature papers setup ─────────────────────────────────────────────────
papers_cfg_file <- file.path(project_root, "config", "papers.csv")
papers_cfg <- tryCatch(
  readr::read_csv(papers_cfg_file, show_col_types = FALSE),
  error = function(e) NULL
)

if (!is.null(papers_cfg)) {
  processed_dir <- file.path(project_root, "data", "processed")

  summary_list <- lapply(papers_cfg$paper_id, function(pid) {
    f <- file.path(processed_dir, paste0(pid, "_staging_summary.csv"))
    if (file.exists(f)) {
      tryCatch(readr::read_csv(f, show_col_types = FALSE),
               error = function(e) NULL)
    } else {
      NULL
    }
  })
  has_summary <- !sapply(summary_list, is.null)

  papers_summary_raw <- if (any(has_summary)) {
    do.call(rbind, summary_list[has_summary])
  } else {
    NULL
  }

  summary_df_for_join <- if (!is.null(papers_summary_raw)) {
    papers_summary_raw
  } else {
    data.frame(
      paper_id = character(0), n_rows = integer(0),
      n_with_coords = integer(0), n_with_country = integer(0),
      n_with_state_province = integer(0), n_with_county_parish = integer(0),
      n_with_municipality = integer(0), n_habit_trait_rows = integer(0),
      n_unique_species = integer(0), generated_at_utc = character(0),
      stringsAsFactors = FALSE
    )
  }

  papers_all <- papers_cfg %>%
    dplyr::left_join(summary_df_for_join, by = "paper_id") %>%
    dplyr::mutate(
      status     = ifelse(!is.na(n_rows), "Processed", "Pending"),
      pct_georef = ifelse(!is.na(n_rows) & n_rows > 0,
                         round(n_with_coords / n_rows * 100, 1), NA_real_)
    )

  # Load georeferenced points from bien_staging.csv files
  lit_georef_list <- lapply(papers_cfg$paper_id, function(pid) {
    f <- file.path(processed_dir, paste0(pid, "_bien_staging.csv"))
    if (!file.exists(f)) return(NULL)
    tryCatch({
      df <- readr::read_csv(f, show_col_types = FALSE)
      if (!all(c("latitude", "longitude") %in% names(df))) return(NULL)
      df <- df[!is.na(df$latitude) & !is.na(df$longitude) &
               df$latitude != "" & df$longitude != "", ]
      if (nrow(df) == 0) return(NULL)
      data.frame(
        paper_id       = pid,
        latitude       = suppressWarnings(as.numeric(df$latitude)),
        longitude      = suppressWarnings(as.numeric(df$longitude)),
        name_submitted = if ("name_submitted" %in% names(df))
                           as.character(df$name_submitted)
                         else NA_character_,
        stringsAsFactors = FALSE
      )
    }, error = function(e) NULL)
  })

  lit_georef_pts <- do.call(rbind, Filter(Negate(is.null), lit_georef_list))
  if (!is.null(lit_georef_pts) && nrow(lit_georef_pts) > 0) {
    lit_georef_pts <- lit_georef_pts[
      !is.na(lit_georef_pts$latitude) & !is.na(lit_georef_pts$longitude), ]
    if (nrow(lit_georef_pts) > 50000) {
      set.seed(42)
      lit_georef_pts <- lit_georef_pts[sample(nrow(lit_georef_pts), 50000), ]
    }
  } else {
    lit_georef_pts <- NULL
  }

  n_papers_total      <- nrow(papers_cfg)
  n_papers_processed  <- sum(has_summary)
  n_papers_pending    <- n_papers_total - n_papers_processed
  n_staging_rows_total <- if (!is.null(papers_summary_raw))
                            sum(papers_summary_raw$n_rows, na.rm = TRUE)
                          else 0L
  n_lit_georef_total  <- if (!is.null(lit_georef_pts)) nrow(lit_georef_pts) else 0L

} else {
  papers_all          <- NULL
  lit_georef_pts      <- NULL
  n_papers_total      <- 0L
  n_papers_processed  <- 0L
  n_papers_pending    <- 0L
  n_staging_rows_total <- 0L
  n_lit_georef_total  <- 0L
}

# ── Consolidated map: combine literature papers + occurrence intake ───────────
# Normalise lit_georef_pts (latitude/longitude/paper_id/name_submitted)
lit_pts_norm <- if (!is.null(lit_georef_pts) && nrow(lit_georef_pts) > 0) {
  data.frame(
    source_label = lit_georef_pts$paper_id,
    lat          = as.numeric(lit_georef_pts$latitude),
    lon          = as.numeric(lit_georef_pts$longitude),
    label        = lit_georef_pts$name_submitted,
    track        = "Literature Paper",
    stringsAsFactors = FALSE
  )
} else NULL

# Normalise all_georef_pts (lat/lon/source_id/species)
occ_pts_norm <- if (!is.null(all_georef_pts) && nrow(all_georef_pts) > 0) {
  data.frame(
    source_label = all_georef_pts$source_id,
    lat          = as.numeric(all_georef_pts$lat),
    lon          = as.numeric(all_georef_pts$lon),
    label        = all_georef_pts$species,
    track        = "Occurrence Intake",
    stringsAsFactors = FALSE
  )
} else NULL

combined_georef_pts <- do.call(rbind, Filter(Negate(is.null), list(lit_pts_norm, occ_pts_norm)))
if (!is.null(combined_georef_pts) && nrow(combined_georef_pts) > 0) {
  combined_georef_pts <- combined_georef_pts[
    !is.na(combined_georef_pts$lat) & !is.na(combined_georef_pts$lon), ]
  if (nrow(combined_georef_pts) > 50000) {
    set.seed(42)
    combined_georef_pts <- combined_georef_pts[sample(nrow(combined_georef_pts), 50000), ]
  }
} else {
  combined_georef_pts <- NULL
}
n_combined_georef <- if (!is.null(combined_georef_pts)) nrow(combined_georef_pts) else 0L
```

# Literature Paper Integration Status

## Executive Sub-Summary

- Total papers configured: **`r n_papers_total`**
- Papers processed/staged: **`r n_papers_processed`**
- Papers pending: **`r n_papers_pending`**
- Total staging rows across all processed papers: **`r n_staging_rows_total`**
- Total georeferenced observations (literature papers): **`r n_lit_georef_total`**

## Per-Paper Status Table

```{r lit-papers-table}
if (!is.null(papers_all)) {
  papers_display <- papers_all %>%
    dplyr::mutate(
      citation_short = substr(citation, 1, 80),
      doi_link = ifelse(
        !is.na(doi) & doi != "",
        sprintf('<a href="https://doi.org/%s" target="_blank">%s</a>', doi, doi),
        "—"
      )
    ) %>%
    dplyr::select(
      paper_id,
      citation_short,
      year,
      journal,
      doi_link,
      status,
      total_records    = n_rows,
      georef_records   = n_with_coords,
      pct_georef,
      unique_species   = n_unique_species,
      habit_trait_rows = n_habit_trait_rows
    )

  DT::datatable(
    papers_display,
    escape     = FALSE,
    rownames   = FALSE,
    options    = list(pageLength = 15, scrollX = TRUE),
    caption    = "All configured literature papers with staging status and quality metrics. Pending papers show NA for numeric fields.",
    colnames   = c(
      "Paper ID", "Citation (truncated)", "Year", "Journal", "DOI",
      "Status", "Total Records", "Georef Records", "% Georef",
      "Unique Species", "Habit/Trait Rows"
    )
  )
} else {
  cat("config/papers.csv not found — literature paper section unavailable.")
}
```

## Geographic Coverage — All Sources

All georeferenced points from both literature papers and occurrence intake sources are shown together in the consolidated map below (Section 3.2).

- Newly compiled intake sources `manual_ukrainian_kurgans_gbif` and `manual_ecat_central_africa` are included when their compiled occurrence files are present.

# Executive Summary

- Report timestamp: `r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`
- Source registry file: `data/occurrence_source_intake.csv`
- Total sources tracked: **`r n_total`**
- Sources accessed/ingested (`harvest_status == "compiled"`): **`r n_accessed`** (`r pct_accessed`%)
- Sources pending (`harvest_status != "compiled"`): **`r n_pending`** (`r pct_pending`%)

## Recent Intake Additions

The following newly compiled intake sources have been added to this report and are now included in the consolidated geographic map.

```{r recent-additions}
if (nrow(recent_additions) > 0) {
  display_additions <- recent_additions %>%
    select(source_id, note_summary, harvest_status)
  knitr::kable(
    display_additions,
    col.names = c("Source ID", "Summary", "Harvest Status"),
    caption = "Recent compiled intake sources included in the report and map"
  )
} else {
  cat("No recent intake additions for the selected source IDs.")
}
```

- `manual_ukrainian_kurgans_gbif`: Moysiyenko et al. 2023 Ukrainian kurgans occurrence dataset, now compiled and mapped.
- `manual_ecat_central_africa`: Tack et al. 2022 ECAT Central Africa expert-validated tree occurrences, now compiled and mapped.

```{r status-breakdown}
knitr::kable(status_counts, caption = "Harvest status breakdown from intake registry")
```

# Sources Already Accessed/Ingested

Evidence columns are included directly from repository artifacts. If evidence is unavailable, fields are marked as `unknown/unverified`.

```{r accessed-table}
accessed_display <- accessed_tbl %>%
  select(
    source_id,
    display_name,
    intake_type,
    source_group,
    canonical_url_or_doi,
    harvest_status,
    compiled_file_exists,
    compiled_row_count,
    notes
  )

DT::datatable(
  accessed_display,
  options = list(pageLength = 15, scrollX = TRUE),
  rownames = FALSE,
  caption = "Sources with compiled status"
)
```

# Compiled Source Quality Summary

Per-source quality metrics computed from compiled occurrence CSVs at render time.

```{r quality-summary-table}
quality_tbl <- intake_enriched %>%
  filter(harvest_status == "compiled") %>%
  select(source_id, display_name,
         total_records    = compiled_row_count,
         georef_records,  pct_georef,
         binomial_records, pct_binomial) %>%
  arrange(source_id)

totals_row <- data.frame(
  source_id        = "TOTAL",
  display_name     = "",
  total_records    = sum(quality_tbl$total_records,    na.rm = TRUE),
  georef_records   = sum(quality_tbl$georef_records,   na.rm = TRUE),
  pct_georef       = round(sum(quality_tbl$georef_records,   na.rm = TRUE) /
                           max(sum(quality_tbl$total_records, na.rm = TRUE), 1) * 100, 1),
  binomial_records = sum(quality_tbl$binomial_records, na.rm = TRUE),
  pct_binomial     = round(sum(quality_tbl$binomial_records, na.rm = TRUE) /
                           max(sum(quality_tbl$total_records, na.rm = TRUE), 1) * 100, 1),
  stringsAsFactors = FALSE
)

quality_display <- rbind(quality_tbl, totals_row)

knitr::kable(
  quality_display,
  col.names = c("Source ID", "Display Name", "Total Records",
                "Georef Records", "% Georef",
                "Binomial Records", "% Binomial"),
  caption = "Per-source quality metrics for compiled occurrence files"
)
```

## Consolidated Geographic Map — All Sources

All **`r n_combined_georef`** georeferenced points across literature papers and compiled occurrence intake sources (sampled to 50,000 max for performance). Points are coloured by source; shape of popup shows data track (Literature Paper vs Occurrence Intake).

- The consolidated map includes the newly ingested `manual_ukrainian_kurgans_gbif` and `manual_ecat_central_africa` sources when their compiled data are available.

```{r consolidated-map}
if (!is.null(combined_georef_pts) && nrow(combined_georef_pts) > 0) {
  all_labels  <- unique(combined_georef_pts$source_label)
  n_labels    <- length(all_labels)
  pal_colors  <- if (n_labels <= 12) {
    RColorBrewer::brewer.pal(max(3L, n_labels), "Paired")[seq_len(n_labels)]
  } else {
    colorRampPalette(RColorBrewer::brewer.pal(12, "Paired"))(n_labels)
  }
  combined_pal <- colorFactor(palette = pal_colors, domain = all_labels)

  leaflet(combined_georef_pts, options = leafletOptions(minZoom = 1)) %>%
    addProviderTiles("CartoDB.Positron") %>%
    addCircleMarkers(
      lng         = ~lon,
      lat         = ~lat,
      color       = ~combined_pal(source_label),
      radius      = 3,
      stroke      = FALSE,
      fillOpacity = 0.65,
      popup       = ~paste0(
        "<b>", track, ":</b> ", source_label,
        "<br><b>Species:</b> ", label
      )
    ) %>%
    addLegend(
      position = "bottomright",
      pal      = combined_pal,
      values   = ~source_label,
      title    = "Source",
      opacity  = 0.85
    )
} else {
  cat("No georeferenced points available to map.")
}
```

# Sources Pending (Waiting List)

```{r pending-table}
pending_display <- pending_tbl %>%
  select(
    source_id,
    display_name,
    intake_type,
    source_group,
    canonical_url_or_doi,
    harvest_status,
    next_action,
    notes
  )

DT::datatable(
  pending_display,
  options = list(pageLength = 20, scrollX = TRUE),
  rownames = FALSE,
  caption = "Sources pending review or manual access"
)
```

# What Has Been Done

```{r done-summary}
knitr::kable(what_done, caption = "Completed (compiled) sources by source_group")
```

- The registry has explicit status entries for all listed sources via `harvest_status`.
- Compiled-source directories were checked under `data/occurrences/<source_id>/compiled_occurrences.csv` for evidence.
- Canonical links/DOIs, action notes, and source notes were pulled from the intake registry without manual reinterpretation.

# What Remains To Be Done

Actionable checklist generated from pending rows:

```{r pending-actions-table}
DT::datatable(
  pending_actions,
  options = list(pageLength = 20, scrollX = TRUE),
  rownames = FALSE,
  caption = "Pending source action checklist"
)
```

- [ ] Resolve all `pending_manual_access` sources with direct coordinator contact and access confirmation.
- [ ] Execute each pending source `next_action` and capture outputs in `data/occurrences/<source_id>/`.
- [ ] Update `harvest_status` to `compiled` only after compiled output is written and validated.
- [ ] Re-render this report after each status update to keep the waiting list current.

# BIEN Species Overlap Analysis

This section compares the species in our literature paper staging data against what is already in the BIEN database, to identify novel taxa not yet represented in BIEN.

**To generate the overlap data:** run the batch query script (requires active BIEN DB connection, ~15–20 min for 66k species):

```bash
cd /Users/brianjenquist/VSCode/Literature_Data_To_BIENdb
Rscript scripts/bien_overlap_check.R
```

The script writes checkpoint files so it can resume if interrupted. Re-run `rmarkdown::render('source_data_progress_report.Rmd')` after the script completes to populate the tables and chart below.

```{r bien-overlap-setup}
bien_detail_file  <- file.path(project_root, "output", "bien_overlap_per_species.csv")
bien_summary_file <- file.path(project_root, "output", "bien_overlap_per_paper.csv")

bien_detail  <- tryCatch(readr::read_csv(bien_detail_file,  show_col_types = FALSE), error = function(e) NULL)
bien_summary <- tryCatch(readr::read_csv(bien_summary_file, show_col_types = FALSE), error = function(e) NULL)

bien_ready <- !is.null(bien_detail) && nrow(bien_detail) > 0 &&
              !is.null(bien_summary) && nrow(bien_summary) > 0
```

```{r bien-exec-summary, results='asis'}
if (bien_ready) {
  n_queried  <- nrow(bien_detail)
  n_in_bien  <- sum(bien_detail$in_bien, na.rm = TRUE)
  n_not_bien <- n_queried - n_in_bien
  pct_in     <- round(100 * n_in_bien / max(n_queried, 1), 1)
  cat(sprintf(
    "- **Unique binomial species queried:** %s\n- **In BIEN:** %s (%s%%)\n- **Not in BIEN (novel):** %s (%s%%)\n",
    format(n_queried, big.mark = ","),
    format(n_in_bien, big.mark = ","), pct_in,
    format(n_not_bien, big.mark = ","), round(100 - pct_in, 1)
  ))
} else {
  cat("> **Results not yet available.** Run `scripts/bien_overlap_check.R` to generate overlap data, then re-render this report.\n")
}
```

## Per-Paper BIEN Overlap Summary

```{r bien-paper-table}
if (bien_ready) {
  DT::datatable(
    bien_summary,
    rownames = FALSE,
    options  = list(pageLength = 15, scrollX = TRUE),
    caption  = "Species overlap with BIEN per literature paper. 'in_bien_count' = species already in BIEN; 'not_in_bien_count' = novel taxa.",
    colnames = c("Paper ID", "Total Species", "In BIEN", "Not in BIEN", "% In BIEN")
  )
} else {
  cat("Run `scripts/bien_overlap_check.R` first.")
}
```

## In-BIEN vs. Novel Species by Paper

```{r bien-bar-chart, fig.width=8, fig.height=4}
if (bien_ready) {
  plot_df <- bien_summary %>%
    tidyr::pivot_longer(
      cols      = c(in_bien_count, not_in_bien_count),
      names_to  = "status",
      values_to = "n_species"
    ) %>%
    dplyr::mutate(
      status = dplyr::recode(status,
        in_bien_count     = "Already in BIEN",
        not_in_bien_count = "Novel (not in BIEN)"
      )
    )

  ggplot2::ggplot(plot_df, ggplot2::aes(x = paper_id, y = n_species, fill = status)) +
    ggplot2::geom_col(position = "stack") +
    ggplot2::coord_flip() +
    ggplot2::scale_fill_manual(
      values = c("Already in BIEN" = "#2166ac", "Novel (not in BIEN)" = "#d73027")
    ) +
    ggplot2::labs(
      x       = NULL,
      y       = "Number of Species",
      fill    = NULL,
      title   = "Species from each literature paper: already in BIEN vs. novel",
      caption = "Novel = species name not matched in BIEN occurrence records."
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(legend.position = "bottom")
} else {
  cat("Run `scripts/bien_overlap_check.R` first.")
}
```

## Top Novel Species (Not in BIEN)

Species absent from BIEN, ranked by how many papers they appear in — highest-priority candidates for BIEN integration.

```{r bien-novel-table}
if (bien_ready) {
  novel_top <- bien_detail %>%
    dplyr::filter(!in_bien) %>%
    dplyr::arrange(dplyr::desc(papers_present_in), species) %>%
    dplyr::slice_head(n = 50) %>%
    dplyr::select(species, bien_record_count, papers_present_in)

  DT::datatable(
    novel_top,
    rownames = FALSE,
    options  = list(pageLength = 25, scrollX = TRUE),
    caption  = "Top 50 species not in BIEN, sorted by number of papers they appear in.",
    colnames = c("Species", "BIEN Records", "Papers Present In")
  )
} else {
  cat("Run `scripts/bien_overlap_check.R` first.")
}
```

# Reproducibility Notes

- This report is fully generated from repository files at render time.
- No source status was inferred from external systems.
- Missing or blank fields are reported as `unknown/unverified`.
- Render command:

```bash
cd /Users/brianjenquist/VSCode/Literature_Data_To_BIENdb
Rscript -e "rmarkdown::render('source_data_progress_report.Rmd')"
```
