---
title: "Literature Data Integration Source Progress Report"
author: "BIEN Literature_Data_To_BIENdb"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(readr)
library(knitr)
library(DT)
library(leaflet)

project_root <- getwd()
intake_file <- file.path(project_root, "data", "occurrence_source_intake.csv")
occurrence_dir <- file.path(project_root, "data", "occurrences")

if (!file.exists(intake_file)) {
  stop("Missing required file: data/occurrence_source_intake.csv")
}

intake <- readr::read_csv(intake_file, show_col_types = FALSE)

required_cols <- c(
  "source_id", "display_name", "intake_type", "source_group",
  "canonical_url_or_doi", "harvest_status", "next_action", "notes"
)
missing_cols <- setdiff(required_cols, names(intake))
if (length(missing_cols) > 0) {
  stop(paste("Missing required columns in intake CSV:", paste(missing_cols, collapse = ", ")))
}

compiled_paths <- file.path(occurrence_dir, intake$source_id, "compiled_occurrences.csv")
compiled_exists <- file.exists(compiled_paths)

read_source_summary <- function(p) {
  tryCatch({
    df <- readr::read_csv(p, show_col_types = FALSE)
    total <- nrow(df)
    georef <- if (all(c("decimalLatitude", "decimalLongitude") %in% names(df))) {
      sum(!is.na(df$decimalLatitude) & !is.na(df$decimalLongitude) &
          df$decimalLatitude != "" & df$decimalLongitude != "", na.rm = TRUE)
    } else { 0L }
    binomial <- if ("species" %in% names(df)) {
      sum(!is.na(df$species) & df$species != "" &
          grepl(" ", trimws(as.character(df$species))), na.rm = TRUE)
    } else { 0L }
    list(total = total, georef = as.integer(georef), binomial = as.integer(binomial), data = df)
  }, error = function(e) {
    list(total = NA_integer_, georef = NA_integer_, binomial = NA_integer_, data = NULL)
  })
}

source_summaries <- vector("list", length(compiled_paths))
for (i in seq_along(compiled_paths)) {
  if (compiled_exists[i]) {
    source_summaries[[i]] <- read_source_summary(compiled_paths[i])
  } else {
    source_summaries[[i]] <- list(total = NA_integer_, georef = NA_integer_, binomial = NA_integer_, data = NULL)
  }
}

compiled_rows     <- sapply(source_summaries, function(s) s$total)
compiled_georef   <- sapply(source_summaries, function(s) s$georef)
compiled_binomial <- sapply(source_summaries, function(s) s$binomial)

intake_enriched <- intake %>%
  mutate(
    canonical_url_or_doi = ifelse(is.na(canonical_url_or_doi) | canonical_url_or_doi == "", "unknown/unverified", canonical_url_or_doi),
    next_action = ifelse(is.na(next_action) | next_action == "", "unknown/unverified", next_action),
    notes = ifelse(is.na(notes) | notes == "", "unknown/unverified", notes),
    compiled_file_exists = ifelse(compiled_exists, "yes", "no"),
    compiled_row_count   = compiled_rows,
    georef_records       = compiled_georef,
    binomial_records     = compiled_binomial,
    pct_georef   = ifelse(!is.na(compiled_rows) & compiled_rows > 0,
                          round(compiled_georef   / compiled_rows * 100, 1), NA_real_),
    pct_binomial = ifelse(!is.na(compiled_rows) & compiled_rows > 0,
                          round(compiled_binomial / compiled_rows * 100, 1), NA_real_),
    status_bucket = case_when(
      harvest_status == "compiled" ~ "accessed_or_ingested",
      harvest_status %in% c("pending_review", "pending_manual_access") ~ "pending",
      TRUE ~ "pending"
    )
  )

# Combined georeferenced points for leaflet map (capped at 50,000 for performance)
all_georef_pts <- do.call(rbind, lapply(seq_along(source_summaries), function(i) {
  s <- source_summaries[[i]]
  if (is.null(s$data)) return(NULL)
  df <- s$data
  if (!all(c("decimalLatitude", "decimalLongitude") %in% names(df))) return(NULL)
  df <- df[!is.na(df$decimalLatitude) & !is.na(df$decimalLongitude) &
           df$decimalLatitude != "" & df$decimalLongitude != "", ]
  if (nrow(df) == 0) return(NULL)
  data.frame(
    source_id = intake$source_id[i],
    species   = if ("species" %in% names(df)) as.character(df$species) else NA_character_,
    lat = suppressWarnings(as.numeric(df$decimalLatitude)),
    lon = suppressWarnings(as.numeric(df$decimalLongitude)),
    stringsAsFactors = FALSE
  )
}))
all_georef_pts <- all_georef_pts[!is.na(all_georef_pts$lat) & !is.na(all_georef_pts$lon), ]
if (!is.null(all_georef_pts) && nrow(all_georef_pts) > 50000) {
  set.seed(42)
  all_georef_pts <- all_georef_pts[sample(nrow(all_georef_pts), 50000), ]
}

accessed_tbl <- intake_enriched %>%
  filter(status_bucket == "accessed_or_ingested") %>%
  arrange(source_group, source_id)

pending_tbl <- intake_enriched %>%
  filter(status_bucket == "pending") %>%
  arrange(harvest_status, source_group, source_id)

status_counts <- intake_enriched %>%
  count(harvest_status, name = "n") %>%
  arrange(desc(n))

n_total <- nrow(intake_enriched)
n_accessed <- nrow(accessed_tbl)
n_pending <- nrow(pending_tbl)
pct_accessed <- ifelse(n_total > 0, round(100 * n_accessed / n_total, 1), NA_real_)
pct_pending <- ifelse(n_total > 0, round(100 * n_pending / n_total, 1), NA_real_)

what_done <- intake_enriched %>%
  filter(status_bucket == "accessed_or_ingested") %>%
  count(source_group, name = "sources_completed") %>%
  arrange(desc(sources_completed), source_group)

pending_actions <- pending_tbl %>%
  transmute(source_id, display_name, harvest_status, next_action)
```

# Executive Summary

- Report timestamp: `r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`
- Source registry file: `data/occurrence_source_intake.csv`
- Total sources tracked: **`r n_total`**
- Sources accessed/ingested (`harvest_status == "compiled"`): **`r n_accessed`** (`r pct_accessed`%)
- Sources pending (`harvest_status != "compiled"`): **`r n_pending`** (`r pct_pending`%)

```{r status-breakdown}
knitr::kable(status_counts, caption = "Harvest status breakdown from intake registry")
```

# Sources Already Accessed/Ingested

Evidence columns are included directly from repository artifacts. If evidence is unavailable, fields are marked as `unknown/unverified`.

```{r accessed-table}
accessed_display <- accessed_tbl %>%
  select(
    source_id,
    display_name,
    intake_type,
    source_group,
    canonical_url_or_doi,
    harvest_status,
    compiled_file_exists,
    compiled_row_count,
    notes
  )

DT::datatable(
  accessed_display,
  options = list(pageLength = 15, scrollX = TRUE),
  rownames = FALSE,
  caption = "Sources with compiled status"
)
```

# Compiled Source Quality Summary

Per-source quality metrics computed from compiled occurrence CSVs at render time.

```{r quality-summary-table}
quality_tbl <- intake_enriched %>%
  filter(harvest_status == "compiled") %>%
  select(source_id, display_name,
         total_records    = compiled_row_count,
         georef_records,  pct_georef,
         binomial_records, pct_binomial) %>%
  arrange(source_id)

totals_row <- data.frame(
  source_id        = "TOTAL",
  display_name     = "",
  total_records    = sum(quality_tbl$total_records,    na.rm = TRUE),
  georef_records   = sum(quality_tbl$georef_records,   na.rm = TRUE),
  pct_georef       = round(sum(quality_tbl$georef_records,   na.rm = TRUE) /
                           max(sum(quality_tbl$total_records, na.rm = TRUE), 1) * 100, 1),
  binomial_records = sum(quality_tbl$binomial_records, na.rm = TRUE),
  pct_binomial     = round(sum(quality_tbl$binomial_records, na.rm = TRUE) /
                           max(sum(quality_tbl$total_records, na.rm = TRUE), 1) * 100, 1),
  stringsAsFactors = FALSE
)

quality_display <- rbind(quality_tbl, totals_row)

knitr::kable(
  quality_display,
  col.names = c("Source ID", "Display Name", "Total Records",
                "Georef Records", "% Georef",
                "Binomial Records", "% Binomial"),
  caption = "Per-source quality metrics for compiled occurrence files"
)
```

## Geographic Distribution of Compiled Occurrences

```{r leaflet-map}
if (!is.null(all_georef_pts) && nrow(all_georef_pts) > 0) {
  pal <- colorFactor(palette = "Set1", domain = unique(all_georef_pts$source_id))

  leaflet(all_georef_pts, options = leafletOptions(minZoom = 1)) %>%
    addTiles() %>%
    addCircleMarkers(
      lng = ~lon, lat = ~lat,
      color = ~pal(source_id),
      radius = 3,
      stroke = FALSE,
      fillOpacity = 0.6,
      popup = ~paste0("<b>Source:</b> ", source_id, "<br><b>Species:</b> ", species)
    ) %>%
    addLegend(
      position = "bottomright",
      pal = pal,
      values = ~source_id,
      title = "Source",
      opacity = 0.8
    )
} else {
  cat("No georeferenced points available to map.")
}
```

# Sources Pending (Waiting List)

```{r pending-table}
pending_display <- pending_tbl %>%
  select(
    source_id,
    display_name,
    intake_type,
    source_group,
    canonical_url_or_doi,
    harvest_status,
    next_action,
    notes
  )

DT::datatable(
  pending_display,
  options = list(pageLength = 20, scrollX = TRUE),
  rownames = FALSE,
  caption = "Sources pending review or manual access"
)
```

# What Has Been Done

```{r done-summary}
knitr::kable(what_done, caption = "Completed (compiled) sources by source_group")
```

- The registry has explicit status entries for all listed sources via `harvest_status`.
- Compiled-source directories were checked under `data/occurrences/<source_id>/compiled_occurrences.csv` for evidence.
- Canonical links/DOIs, action notes, and source notes were pulled from the intake registry without manual reinterpretation.

# What Remains To Be Done

Actionable checklist generated from pending rows:

```{r pending-actions-table}
DT::datatable(
  pending_actions,
  options = list(pageLength = 20, scrollX = TRUE),
  rownames = FALSE,
  caption = "Pending source action checklist"
)
```

- [ ] Resolve all `pending_manual_access` sources with direct coordinator contact and access confirmation.
- [ ] Execute each pending source `next_action` and capture outputs in `data/occurrences/<source_id>/`.
- [ ] Update `harvest_status` to `compiled` only after compiled output is written and validated.
- [ ] Re-render this report after each status update to keep the waiting list current.

# Reproducibility Notes

- This report is fully generated from repository files at render time.
- No source status was inferred from external systems.
- Missing or blank fields are reported as `unknown/unverified`.
- Render command:

```bash
cd /Users/brianjenquist/VSCode/Literature_Data_To_BIENdb
Rscript -e "rmarkdown::render('source_data_progress_report.Rmd')"
```
