---
title: "Literature-Sourced Occurrence Data: Quality & Geographic Overview"
author: "BIEN Database Project"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    code_folding: hide
params:
  data_dir: "data/processed"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 150
)

# Load required packages
library(dplyr)
library(tidyr)
library(knitr)
library(leaflet)
library(sf)
library(DT)
library(ggplot2)

# Define project root
project_root <- here::here()
data_dir <- file.path(project_root, params$data_dir)
```

```{r load_metadata}
# Load paper configuration
papers_config <- read.csv(
  file.path(project_root, "config/papers.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# Create mapping of paper_id to citation
paper_metadata <- papers_config %>%
  select(paper_id, citation, doi, year, publisher, journal, notes)

# Extract processed papers (those with _bien_staging.csv files)
processed_files <- list.files(
  data_dir,
  pattern = "_bien_staging\\.csv$",
  full.names = FALSE
)

processed_paper_ids <- gsub("_bien_staging\\.csv$", "", processed_files)

cat("Found", length(processed_paper_ids), "processed papers\n")
cat("Papers:", paste(processed_paper_ids, collapse = ", "), "\n")
```

```{r load_all_data}
# Function to load and label data with paper_id
load_paper_data <- function(paper_id, data_dir) {
  file_path <- file.path(data_dir, paste0(paper_id, "_bien_staging.csv"))
  
  if (!file.exists(file_path)) {
    return(NULL)
  }
  
  d <- read.csv(
    file_path,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
  
  # Add paper_id if not present
  if (!"paper_id" %in% names(d)) {
    d$paper_id <- paper_id
  }
  
  return(d)
}

# Load all papers
all_data <- list()
for (pid in processed_paper_ids) {
  dat <- load_paper_data(pid, data_dir)
  if (!is.null(dat)) {
    # Standardize common columns
    if ("occurrence_id" %in% names(dat)) {
      dat$occurrence_id <- as.character(dat$occurrence_id)
    }
    all_data[[pid]] <- dat
  }
}

# Combine into single dataframe
combined_data <- bind_rows(all_data, .id = NULL)

cat("Total rows loaded:", nrow(combined_data), "\n")
cat("Unique papers:", n_distinct(combined_data$paper_id), "\n")
```

```{r calculate_quality_metrics}
# Calculate quality metrics
quality_metrics <- combined_data %>%
  group_by(paper_id) %>%
  summarise(
    total_records = n(),
    georeferenced = sum(!is.na(decimalLatitude) & !is.na(decimalLongitude)),
    with_taxon = sum(!is.na(name_submitted) & nzchar(name_submitted)),
    with_date = sum(!is.na(date_collected) & nzchar(date_collected)),
    with_country = sum(!is.na(country) & nzchar(country)),
    distinct_species = n_distinct(name_submitted, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pct_georeferenced = round(100 * georeferenced / total_records, 1),
    pct_with_taxon = round(100 * with_taxon / total_records, 1),
    pct_with_date = round(100 * with_date / total_records, 1)
  )

# Add paper metadata
quality_metrics <- quality_metrics %>%
  left_join(
    paper_metadata %>% select(paper_id, citation, year, publisher),
    by = "paper_id"
  ) %>%
  select(paper_id, year, publisher, total_records, georeferenced, pct_georeferenced,
         with_taxon, pct_with_taxon, distinct_species, citation)

# Calculate totals
total_records <- sum(quality_metrics$total_records)
total_georeferenced <- sum(quality_metrics$georeferenced)
total_species <- n_distinct(combined_data$name_submitted, na.rm = TRUE)
```

# Executive Summary

This report provides a comprehensive overview of occurrence data loaded from **`r length(processed_paper_ids)` peer-reviewed literature sources** into the BIEN database pipeline.

## Key Statistics

- **Total observations:** `r format(total_records, big.mark=",")`
- **Georeferenced observations:** `r format(total_georeferenced, big.mark=",")` (`r round(100*total_georeferenced/total_records, 1)`%)
- **Distinct species:** `r format(total_species, big.mark=",")`
- **Papers processed:** `r length(processed_paper_ids)`

---

# Data Quality by Paper

```{r quality_table}
quality_display <- quality_metrics %>%
  select(
    paper_id,
    year,
    total_records,
    georeferenced,
    pct_georeferenced,
    with_taxon,
    pct_with_taxon,
    distinct_species
  ) %>%
  arrange(desc(total_records))

datatable(
  quality_display,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = c('csv', 'excel', 'pdf'),
    columnDefs = list(
      list(width = '120px', targets = 0),
      list(width = '80px', targets = c(1, 2, 3, 4, 5, 6, 7))
    )
  ),
  colnames = c(
    'Paper ID',
    'Year',
    'Total Records',
    'Georeferenced',
    '% Geo',
    'With Taxon',
    '% Taxon',
    'Distinct Species'
  ),
  rownames = FALSE
)
```

---

# Geographic Coverage

## Interactive Map

The map below shows all georeferenced observation points from the literature sources. Hover over points to see the taxon name and source paper.

```{r prepare_map_data}
# Prepare data for mapping
map_data <- combined_data %>%
  mutate(
    map_lat = suppressWarnings(as.numeric(decimalLatitude)),
    map_lon = suppressWarnings(as.numeric(decimalLongitude))
  ) %>%
  filter(
    !is.na(map_lat),
    !is.na(map_lon),
    map_lat >= -90,
    map_lat <= 90,
    map_lon >= -180,
    map_lon <= 180
  ) %>%
  select(
    paper_id,
    name_submitted,
    map_lat,
    map_lon,
    country,
    date_collected
  ) %>%
  mutate(
    name_submitted = ifelse(is.na(name_submitted) | !nzchar(name_submitted),
                            "Unknown species", name_submitted),
    popup_text = paste0(
      "<strong>", name_submitted, "</strong><br>",
      "Paper: ", paper_id, "<br>",
      "Country: ", ifelse(is.na(country) | !nzchar(country), "Unknown", country), "<br>",
      "Date: ", ifelse(is.na(date_collected) | !nzchar(date_collected), "Unknown", date_collected)
    ),
    popup_hover = paste0(
      "<strong>", name_submitted, "</strong><br>",
      paper_id
    )
  )

cat("Map points prepared:", nrow(map_data), "\n")
cat("Papers with map points:", n_distinct(map_data$paper_id), "\n")

# Define color palette for papers
paper_colors <- setNames(
  grDevices::hcl.colors(length(processed_paper_ids), palette = "Set 2"),
  processed_paper_ids
)
```

```{r create_interactive_map, fig.width=14, fig.height=10}
# Create base map
m <- leaflet(map_data, 
             options = leafletOptions(
               maxBounds = list(
                 list(-90, -180),
                 list(90, 180)
               )
             )) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = 0, lat = 20, zoom = 2)

# Add layer for each paper
for (paper in processed_paper_ids) {
  paper_data <- map_data %>% filter(paper_id == paper)
  
  if (nrow(paper_data) > 0) {
    m <- m %>%
      addCircleMarkers(
        data = paper_data,
        lng = ~map_lon,
        lat = ~map_lat,
        radius = 5,
        color = paper_colors[paper],
        stroke = TRUE,
        weight = 1,
        opacity = 0.7,
        fill = TRUE,
        fillColor = paper_colors[paper],
        fillOpacity = 0.6,
        popup = ~popup_text,
        label = ~popup_hover,
        labelOptions = labelOptions(style = list("font-size" = "12px")),
        group = paper
      )
  }
}

# Add layer control
m <- m %>%
  addLayersControl(
    overlayGroups = processed_paper_ids,
    options = layersControlOptions(collapsed = FALSE)
  )

# Print map
m
```

### Map Legend

Each paper is represented by a distinct color:

```{r map_legend_table}
legend_df <- data.frame(
  Paper = names(paper_colors),
  Color = as.character(paper_colors),
  stringsAsFactors = FALSE
) %>%
  left_join(
    paper_metadata %>% select(paper_id, year, journal),
    by = c("Paper" = "paper_id")
  )

kable(
  legend_df,
  col.names = c("Paper ID", "Color", "Year", "Journal"),
  format = "html",
  escape = FALSE
)
```

---

# Geographic Distribution Summary

```{r geographic_summary}
# Count observations by country
country_summary <- combined_data %>%
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) %>%
  group_by(country) %>%
  summarise(
    observations = n(),
    papers = n_distinct(paper_id),
    species = n_distinct(name_submitted, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(observations)) %>%
  head(20)

kable(
  country_summary,
  col.names = c("Country", "Observations", "Papers", "Species"),
  format = "html",
  caption = "Top 20 Countries by Observation Count"
)
```

---

# Temporal Coverage

```{r temporal_analysis}
# Extract year from date_collected
temporal_data <- combined_data %>%
  filter(!is.na(date_collected)) %>%
  mutate(
    year = as.numeric(substr(date_collected, 1, 4))
  ) %>%
  filter(!is.na(year) & year > 1500 & year < 2030)

year_range <- paste0(
  "Year range: ",
  min(temporal_data$year, na.rm = TRUE),
  " – ",
  max(temporal_data$year, na.rm = TRUE)
)
```

`r year_range`

```{r temporal_plot, fig.width=12, fig.height=5}
# Plot temporal distribution by paper
temporal_by_paper <- temporal_data %>%
  group_by(paper_id, year) %>%
  summarise(count = n(), .groups = "drop")

ggplot(temporal_by_paper, aes(x = year, y = count, fill = paper_id)) +
  geom_col(position = "stack") +
  facet_wrap(~paper_id, scales = "free_y", nrow = 3) +
  labs(
    title = "Temporal Distribution of Observations by Paper",
    x = "Year",
    y = "Number of Observations",
    fill = "Paper"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# Data Completeness

```{r completeness_analysis}
# Calculate completeness metrics
completeness <- combined_data %>%
  summarise(
    total_records = n(),
    with_taxon = sum(!is.na(name_submitted) & nzchar(name_submitted)),
    with_lat_lon = sum(!is.na(decimalLatitude) & !is.na(decimalLongitude)),
    with_date = sum(!is.na(date_collected) & nzchar(date_collected)),
    with_country = sum(!is.na(country) & nzchar(country)),
    with_elevation = sum(!is.na(elevation_m)),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = starts_with("with_"),
    names_to = "field",
    values_to = "count"
  ) %>%
  mutate(
    field = gsub("^with_", "", field),
    percentage = round(100 * count / total_records, 1)
  ) %>%
  select(field, count, percentage) %>%
  arrange(desc(percentage))

kable(
  completeness,
  col.names = c("Field", "Records with Data", "% Complete"),
  format = "html",
  caption = "Overall Data Completeness"
)
```

```{r completeness_plot, fig.width=12, fig.height=5}
ggplot(completeness, aes(x = reorder(field, percentage), y = percentage)) +
  geom_col(fill = "#2E86AB", alpha = 0.8) +
  geom_text(aes(label = paste0(percentage, "%")), vjust = -0.3, hjust = 0.5) +
  coord_flip() +
  labs(
    title = "Data Field Completeness Across All Papers",
    x = "Field",
    y = "% Records with Data"
  ) +
  theme_minimal() +
  ylim(0, 105)
```

---

# Taxonomic Diversity

```{r taxonomic_analysis}
# Top species across all data
top_species <- combined_data %>%
  filter(!is.na(name_submitted) & nzchar(name_submitted)) %>%
  group_by(name_submitted) %>%
  summarise(
    observations = n(),
    papers = n_distinct(paper_id),
    countries = n_distinct(country, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(observations)) %>%
  head(30)

datatable(
  top_species,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = c('csv', 'excel'),
    pageLength = 15
  ),
  colnames = c('Species', 'Observations', 'Papers', 'Countries'),
  rownames = FALSE,
  caption = "Top 30 Most Frequently Observed Species"
)
```

---

# Data Quality Across Papers

```{r quality_by_paper_detailed}
# Detailed quality metrics by paper
quality_by_paper <- combined_data %>%
  group_by(paper_id) %>%
  summarise(
    total_obs = n(),
    has_coords = sum(!is.na(decimalLatitude) & !is.na(decimalLongitude)),
    has_taxon = sum(!is.na(name_submitted) & nzchar(name_submitted)),
    has_date = sum(!is.na(date_collected) & nzchar(date_collected)),
    has_country = sum(!is.na(country) & nzchar(country)),
    has_elevation = sum(!is.na(elevation_m)),
    distinct_species = n_distinct(name_submitted, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pct_coords = round(100 * has_coords / total_obs, 1),
    pct_taxon = round(100 * has_taxon / total_obs, 1),
    pct_date = round(100 * has_date / total_obs, 1)
  ) %>%
  left_join(
    paper_metadata %>% select(paper_id, citation, year),
    by = "paper_id"
  ) %>%
  select(paper_id, year, total_obs, has_coords, pct_coords, has_taxon,
         pct_taxon, has_date, pct_date, distinct_species)

kable(
  quality_by_paper,
  col.names = c("Paper", "Year", "Total", "Coords", "%", "Taxon", "%", 
                "Date", "%", "Species"),
  format = "html",
  caption = "Detailed Quality Metrics by Paper"
)
```

---

# Session Information

```{r session_info}
sessionInfo()
```

**Report generated:** `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`
